\chapter{Limitations and restrictions}\label{limits}
\index{implementation limitations}
\index{limitations of implementation}

There are some limitations and restrictions in implementation of both
\mt{} and \ot{} compilers.

\section*{Length of identifiers}

The length of an identifier is at most 127 characters.

\ifgenc
There may be further limitations imposed by your C compiler and linker.
Please check the appropriate reference manuals.
\fi % \ifgenc

\section*{Length of literal strings}

The  length of a literal string is at most 256 characters. Longer
strings  may  be  constructed  using the string concatenation
operator (See \ref{m2:ISO:strings}).

\section*{Record extension hierarchy}

The depth of a record extension hierarchy is at most 15 extensions.

\section*{Unimplemented ISO libraries}

\ifgenc
The following \mt{} ISO standard library modules are not
available in the current release:
\begin{flushleft}
\begin{tabular}{ll}
\bf TermFile & Access to an interactive terminal                      \\
\bf LowLong  & Access to underlying properties of the type {\tt LONGREAL}   \\
\bf LowReal  & Access to underlying properties of the type {\tt REAL}   \\
\end{tabular}
\end{flushleft}
\lindex{TermFile}
\lindex{LowLong}
\lindex{LowReal}
\fi % \ifgencode

\section*{Unimplemented Oakwood libraries}

The following \ot{} Oakwood library modules are not
available in the current release:
\begin{flushleft}
\begin{tabular}{lp{7cm}}
\bf Input   & Keyboard and pointer device access \\
\bf Files   & File input/output, riders          \\
\bf XYPlane & Elementary pixel plotting          \\
\end{tabular}
\end{flushleft}
\lindex{Input}
\lindex{Files}
\lindex{XYPlane}

\ifgenc
\section*{Whole overflow}

The whole overflow exception is not detected in the current release.
\fi % \ifgencode

\section*{Coroutines}

The current release provides a restricted implementation of the
system module {\bf COROUTINES}: the interrupt requests are not
detected.

\section*{Dynamic loader}

The \ot{} dynamic loading facility is not provided in the current release.
