\newif\ifonline

\documentstyle[11pt,makeidx,a4]{book}

\title{LaTeX OnLine Users' Guide}

\newcommand{\LO}{LaTeX OnLine}

\setlength{\parindent}{0 pt}
\setlength{\parskip}{6 pt}

\begin{document}

\noindent

\chapter{About LaTeX OnLine}

{\bf \LO} is a translator from a subset of \LaTeX{}
macro package to various on-line document formats, either
target or intermediate. The currently supported formats are:

\begin{itemize}
\item IPF tag language, which is a source for OS/2 INF and
      HLP files. An IPF compiler is available as a part of IBM
      Developer's Toolkit and along with some C/C++ compilers.
\item Microsoft Help Compiler compatible subset of RTF,
      which can be used to produce a Windows Help file.
\item HTML, which can be viewed with any Web browser
      immediately after translation.
\end{itemize}

\LO{} was originally developed for xTech internal use.
The current manual is very brief and was written with assumption
that a reader is used to \LaTeX{}. Its source, is bundled with the
program to serve as a sample.

\chapter{Using \LO{}}

To invoke \LO{}, type

\verb'    convert {<option>} <source file>'

at the command prompt and press Enter. An option is
preceded with either slash (\verb'"/"') or minus sign.
Available options are:

\begin{center}
\begin{tabular}{ll}
\verb'-i'    & emit IPF   \\
\verb'-r[c[#]]' & emit RTF and, optionally, CNT (\verb'#' is a top heading level) \\
\verb'-h'    & emit HTML  \\
\verb'-e'    & emit extern file (see \ref{})
\end{tabular}
\end{center}

In the current version, if no output format is specified, IPF is assumed.

On startup, \verb'convert' searches for a file called \verb'convert.ini'
in the current directory and then in the directory where the executable
resides. This file consists of lines in format:

\verb'<LaTeX command>'              \\
\verb'<optional arguments number>'  \\
\verb'<mandatory arguments number>'

The commands listed in this file will be ignored along with their arguments.
Any commands. which should have no effect on on-line documents anyway,
may be added to it.

Format of \verb'convert.ini' is subject to change in future versions of
\LO. It will contain configuration options for all output formats.

\verb'convert' reports warnings and errors in format:

\verb'"(" <file> <line> ":" <column> ")" <message>'

Processing is stopped on first encountered error.

\chapter{Supported LaTeX features}

The structure of this chapter closely matches the Appendix C of Leslie
Lamport's book called \LaTeX: {\em A Document Preparation System}, which
was used as a main information source during the development of \LO.
With rare exceptions, the supported features exist in \LaTeX{} 2.09.
They are is very briefly listed in the following sections.

\section{Commands and Environments}

\verb'    {}' (scopes) \\
\verb'    \begin{'{\it name}\verb'} ... \end{'{\it name}\verb'}'
\verb'    \\'

The *-form and optional argument are not supported for \verb'\\'.

\section{Sentences and Paragraphs}

\subsection{Making Sentences}

{\bf special characters}

\verb'    \$ \% \{ \_ \& \# \}'

{\bf logos}

\verb'    \TeX \LaTeX'

\subsection{Making Paragraphs}

\verb'    \par'

The \verb'\indent' and \verb'\noindent' commands are currently not supported.
In produced output, paragraphs are always indentless.

Style parameters are not supported. % !!!

\subsection{Footnotes}

\verb'    \footnote{'{\it text}\verb'}'

Optional \verb'\footnote' argument is not supported.

Style parameters are not supported. % !!!

\subsection{Accents and Special Symbols}

Accents are not supported.

Many special, foreign, and  mathematical symbols are recognized,
but only a few of them are supported in output formats.

\section{Sectioning and Table of Contents}

\subsection{Sectioning Commands}

\verb'    \part'          \\
\verb'    \chapter'       \\
\verb'    \section'       \\
\verb'    \subsection'    \\
\verb'    \subsubsection' \\
\verb'    \paragraph'     \\
\verb'    \subparagraph'

*-forms are also supported.

The optional {\it toc\_entry} argument is not supported yet.

{\bf Warning:} the \verb'\part' command is recognized, but its usage may
cause unpredictable results in the current version.

The \verb'\appendix' declaration is ignored.

\subsection{Table of Contents}

A table of contents is always formed, but its placement and representation
depends on the target format. % !!! desribe.

\subsection{Style parameters}

Section numbering and table of contents style parameters ar not supported.

\section{Document and Page Styles}

\verb'    \title{'{\it text}\verb'}'

The only command from this group which is supported is \verb'\title'.
It defines the on-line document title.

\section{Displayed paragraphs}

\subsection{Quotations and Verse}

Nothing is supported.

\subsection{Horizontal Alignment}

\verb'    \begin{flushleft}  ...  \end{flushleft}'  \\
\verb'    \begin{flushright} ...  \end{flushright}' \\
\verb'    \begin{center}     ...  \end{center}'

The \verb'flushleft' environment has no effect since output is left
flushed by default.

{\bf Note:} \verb'flushright' and \verb'center' environments are not
supported in IPF.

\subsection{List-making Environments}

\verb'    \begin{itemize}     '{\it item\_list}\verb' \end{itemize}'     \\
\verb'    \begin{enumerate}   '{\it item\_list}\verb' \end{enumerate}'   \\
\verb'    \begin{description} '{\it item\_list}\verb' \end{description}' \\
\verb'    \item['{\it label}\verb']'

The \verb'list' and \verb'trivlist' environments are not supported.

\subsection{Verbatim}

\verb'    \begin{verbatim} '{\it literal\_text}\verb' \end{verbatim}'  \\
\verb'    \verb'{\it char literal\_text char}

*-forms are not supported.

\section{Mathematical formulas}

In the current version, the one-character command \verb'$' causes font
to be switched to roman style. No other math mode features are supported.

\section{Definitions}

\verb'    \newcommand  {'{\it cmd}\verb'}['{\it args}\verb']{'{\it def}\verb'}' \\
\verb'    \renewcommand{'{\it cmd}\verb'}['{\it args}\verb']{'{\it def}\verb'}'

\verb'#'{\it n} parameters are supported. Arguments and {\it def} may contain
other user-defined commands.

\section{Figures and Other Floating bodies}

\verb'    \begin{table}['{\it loc}\verb'] '{\it body}\verb' \end{table}'

The *-form and the \verb'figure' environment are not supported yet.

\verb'    \caption{'{\it heading}\verb'}'

Optional {\it lst\_entry} parameter of the \verb'\caption' command is not
supported. The \verb'\label' command should {\em not} be placed inside
{\it heading}.

Style parameters are not supported.

Marginal notes are not supported.

\section{Lining It Up in Columns}

\subsection{The tabbing Environment}

Not supported yet.

\subsection{The array and tabular Environments}

\verb'    \begin{tabular}['{\it pos}\verb']{'{\it cols}\verb'} '{\it body}\verb' \end{table}'

The \verb'array' and \verb'tabular*' environments are not supported yet.

Tables is the most difficult thing to translate, so the results
shown by the current version most probably will not satisfy you.

\section{Moving information around}

\subsection{Cross-References}

\verb'    \label{'{\it key}\verb'}'  \\
\verb'    \ref{'{\it key}\verb'}'    \\

%\verb'    \public'                   \\ !!!
%\verb'    \extern'                      !!!

\subsection{Bibliography and Citation}

Not supported yet.

\subsection{Splitting the input}

\verb'    \input{'{\it file\_name}\verb'}'   \\
\verb'    \include{'{\it file}\verb'}'

The \verb'\includeonly' command is not currently supported.

\subsection{Index}

\verb'    \index{'{\it str}\verb'}'

Index is not supported for HTML yet. The format of {\it str} is
compatible with \verb'makeindx' 2.12, but is not fully supported yet.

\section{Line and Page Breaking}

\verb'    \\'            \\
\verb'    \newline'

The optional {\it len} argument of \verb'\\' is not supported.

\section{Length, Spaces, and Boxes}

Nothing is supported.

% \verb'    \mbox        break;                     // mbox == group !!!'

\section{The picture Environment}

Not supported.

\section{Font Selection}

\subsection{Changing the Type Style}

The following type style selection declarations are recognized:

\verb'    \rm \bf \sf \it \sl \sc \tt \em'

However, for a particular output format some of them may be mapped
to the same type style.

\subsection{Changing the Type Size}

The following font size selection declarations are recognized:

\begin{verbatim}
     \tiny        \scriptsize   \footnotesize
     \small       \normalsize   \large
     \Large       \LARGE        \huge
     \Huge
\end{verbatim}

However, for a particular output format some of them may be mapped
to the same size.

\section{Conditional sections}

\verb'    \newif\if'{\it name} \\
\verb'    \if'{\it name}    \\
\verb'    \else'  \\
\verb'    \fi'    \\
\verb'    \'{\it name}\verb'true'  \\
\verb'    \'{\it name}\verb'false'

In \LO, there is a special value of {\it name}:
\verb'\newif\ifonline' implies \verb'\onlinetrue'
and \verb'\onlinefalse' has no effect.

This is not a LaTeX 2.09 feature.

% !!! \verb'    \output' drop it off

\end{document}


